---
title: "HLM_TST_Student_role_only"
author: "Clara Belitz"
date: "11/23/2022"
output: html_document
---


```{r setup, include=FALSE}
library(lme4)      # for fitting model
library(lmerTest)  # for getting df, t and p for fixed effects
library(texreg)    # nice formated output (laTex included)
library(optimx)    # needed for changing algorithm
```

## Data Setup

We need to load the general training data, make student ID and student role factor variables (with "none" as the default level), and then create a new variable called "Correct" that notes whether the model was correct for a given prediction. The models are random forest, decision tree, neural net, XGBoost, and a dummy model, as defined in the previous detector rot paper.

```{r data-setup}
# load training data
detector_rot_bias_data <- read.csv("data/category_ethnicity_race.csv", header=TRUE)
head(detector_rot_bias_data)


#make student id and categories of interest factor variables
detector_rot_bias_data$StudentID<-as.factor(detector_rot_bias_data$student_id)
detector_rot_bias_data$role = factor(detector_rot_bias_data$role, levels=c("None", "negative", "neutral", "positive")) # none is baseline level

# create column to note whether a given observation was predicted correctly or not, for each model
detector_rot_bias_data$rf_correct<- with(detector_rot_bias_data, ifelse(rf_pred == target, 1, 0))
detector_rot_bias_data$dt_correct <- with(detector_rot_bias_data, ifelse(dt_pred == target, 1, 0))
detector_rot_bias_data$nn_correct <- with(detector_rot_bias_data, ifelse(nn_pred == target, 1, 0))
detector_rot_bias_data$xgb_correct <- with(detector_rot_bias_data, ifelse(xgb_pred == target, 1, 0))
detector_rot_bias_data$dm_correct <- with(detector_rot_bias_data, ifelse(dm_pred == target, 1, 0))

```
Student role is mentioned quite often, with only 154 out of 458 responses not containing references to this category. Of the responses, the vast majority are neutral (241).

``` {r response-counts, echo=FALSE}
table(detector_rot_bias_data$role)
```

## Basic HLM explaining whether the model was correct for each model
###Analogous to Overall Accuracy Equality.

```{r correct, echo=FALSE}
rf_model_role_aoe <- glmer(rf_correct ~ role + (1 | StudentID), data=detector_rot_bias_data,
                   family=binomial)
summary(rf_model_role_aoe)


dt_model_role_aoe <- glmer(dt_correct ~ role + (1 | StudentID), data=detector_rot_bias_data,
                   family=binomial)
summary(dt_model_role_aoe)

nn_model_role_aoe <- glmer(nn_correct ~ role + (1 | StudentID), data=detector_rot_bias_data,
                   family=binomial)
summary(nn_model_role_aoe)

xgb_model_role_aoe <- glmer(xgb_correct ~ role + (1 | StudentID), data=detector_rot_bias_data,
                    family=binomial)
summary(xgb_model_role_aoe)

dummy_model_role_aoe <- glmer(dm_correct ~ role + (1 | StudentID), data=detector_rot_bias_data,
                      family=binomial)
summary(dummy_model_role_aoe)
```

The models demonstrate significance for some of the values, generally demonstrating that students who mentioned positive student roles are less likely to have correct predictions. That is, the model is incorrect more often. Thought the coeffecients for negative valence look dramatic, the z-scores are extremely small, likely reflecting the small group size, which makes it difficult to draw conclusions.

### Analogous to Statistical Parity
We can try the models again looking at the prediction of positives only and negatives only, rather than whether the model was correct in general. This is analogous to statistical parity. Because the positive and negative models would be equivalent but reversed we can run the model just once.

```{r stat-parity, echo=FALSE}
rf_model_role_sp <- glmer(rf_pred ~ role + (1 | StudentID), data=detector_rot_bias_data,
                   family=binomial)
summary(rf_model_role_sp)


dt_model_role_sp <- glmer(dt_pred ~ role + (1 | StudentID), data=detector_rot_bias_data,
                   family=binomial)
summary(dt_model_role_sp)

nn_model_role_sp <- glmer(nn_pred ~ role + (1 | StudentID), data=detector_rot_bias_data,
                   family=binomial)
summary(nn_model_role_sp)

xgb_model_role_sp <- glmer(xgb_pred ~ role + (1 | StudentID), data=detector_rot_bias_data,
                    family=binomial)
summary(xgb_model_role_sp)

dummy_model_role_sp <- glmer(dm_pred ~ role + (1 | StudentID), data=detector_rot_bias_data,
                      family=binomial)
summary(dummy_model_role_sp)
```

With this data, role does not appear to affect the prediction rates. 

### Analogous to conditional procedure accuracy (break groups into pos and neg labels)
We can look at conditional procedure accuracy with two HLMs per machine learning model. We first break the datasets into two groups based on the ground truth labels: observations where gaming was occurring and observations were gaming was not occurring.

We then look at likelihood of correct predictions for each role group within the subset.
```{r cpa, echo=FALSE}
gaming_true_labels = detector_rot_bias_data[detector_rot_bias_data$target == 1, ]

gaming_false_labels = detector_rot_bias_data[detector_rot_bias_data$target == 0, ]



rf_true_model_role_cpa <- glmer(rf_correct ~ role + (1 | StudentID), data=gaming_true_labels,
                            family=binomial)
summary(rf_true_model_role_cpa)

rf_false_model_role_cpa <- glmer(rf_correct ~ role + (1 | StudentID), data=gaming_false_labels,
                            family=binomial)
summary(rf_false_model_role_cpa)
# -------------------------------

dt_true_model_role_cpa <- glmer(dt_correct ~ role + (1 | StudentID), data=gaming_true_labels,
                            family=binomial)
summary(dt_true_model_role_cpa)

dt_false_model_role_cpa <- glmer(dt_correct ~ role + (1 | StudentID), data=gaming_false_labels,
                            family=binomial)
summary(dt_false_model_role_cpa)

# -------------------------------

nn_true_model_role_cpa <- glmer(nn_correct ~ role + (1 | StudentID), data=gaming_true_labels,
                            family=binomial)
summary(nn_true_model_role_cpa)

nn_false_model_role_cpa <- glmer(nn_correct ~ role + (1 | StudentID), data=gaming_false_labels,
                            family=binomial)
summary(nn_false_model_role_cpa)

# -------------------------------

xgb_true_model_role_cpa <- glmer(xgb_correct ~ role + (1 | StudentID), data=gaming_true_labels,
                            family=binomial)
summary(xgb_true_model_role_cpa)

xgb_false_model_role_cpa <- glmer(xgb_correct ~ role + (1 | StudentID), data=gaming_false_labels,
                             family=binomial)
summary(xgb_false_model_role_cpa)
# -------------------------------

dm_true_model_role_cpa <- glmer(dm_correct ~ role + (1 | StudentID), data=gaming_true_labels,
                            family=binomial)
summary(dm_true_model_role_cpa)

dm_false_model_role_cpa <- glmer(dm_correct ~ role + (1 | StudentID), data=gaming_false_labels,
                             family=binomial)
summary(dm_false_model_role_cpa)
```


### Analogous to conditional use accuracy (break groups into pos and neg predictions)
We can also look at conditional use accuracy with two HLMs per machine learning model. We first break the datasets into two groups based on the predicted labels: observations where gaming was predicted as occurring and observations were gaming was not predicted.

We then look at likelihood of correct predictions for each role group within the subset.
```{r cua, echo=FALSE}
rf_predicted_gaming = detector_rot_bias_data[detector_rot_bias_data$rf_pred == 1, ]
dt_predicted_gaming = detector_rot_bias_data[detector_rot_bias_data$dt_pred == 1, ]
nn_predicted_gaming = detector_rot_bias_data[detector_rot_bias_data$nn_pred == 1, ]
xgb_predicted_gaming = detector_rot_bias_data[detector_rot_bias_data$xgb_pred == 1, ]
dummy_predicted_gaming = detector_rot_bias_data[detector_rot_bias_data$dm_pred == 1, ]

rf_predicted_not_gaming = detector_rot_bias_data[detector_rot_bias_data$rf_pred == 0, ]
dt_predicted_not_gaming = detector_rot_bias_data[detector_rot_bias_data$dt_pred == 0, ]
nn_predicted_not_gaming = detector_rot_bias_data[detector_rot_bias_data$nn_pred == 0, ]
xgb_predicted_not_gaming = detector_rot_bias_data[detector_rot_bias_data$xgb_pred == 0, ]
dummy_predicted_not_gaming = detector_rot_bias_data[detector_rot_bias_data$dm_pred == 0, ]


rf_true_model_role_cua <- glmer(rf_correct ~ role + (1 | StudentID), data=rf_predicted_gaming,
                            family=binomial)
summary(rf_true_model_role_cua)

rf_false_model_role_cua <- glmer(rf_correct ~ role + (1 | StudentID), data=rf_predicted_not_gaming,
                             family=binomial)
summary(rf_false_model_role_cua)
# -------------------------------

dt_true_model_role_cua <- glmer(dt_correct ~ role + (1 | StudentID), data=dt_predicted_gaming,
                            family=binomial)
summary(dt_true_model_role_cua)

dt_false_model_role_cua <- glmer(dt_correct ~ role + (1 | StudentID), data=dt_predicted_not_gaming,
                             family=binomial)
summary(dt_false_model_role_cua)

# -------------------------------

nn_true_model_role_cua <- glmer(nn_correct ~ role + (1 | StudentID), data=nn_predicted_gaming,
                            family=binomial)
summary(nn_true_model_role_cua)

nn_false_model_role_cua <- glmer(nn_correct ~ role + (1 | StudentID), data=nn_predicted_not_gaming,
                             family=binomial)
summary(nn_false_model_role_cua)

# -------------------------------

xgb_true_model_role_cua <- glmer(xgb_correct ~ role + (1 | StudentID), data=xgb_predicted_gaming,
                             family=binomial)
summary(xgb_true_model_role_cua)

xgb_false_model_role_cua <- glmer(xgb_correct ~ role + (1 | StudentID), data=xgb_predicted_not_gaming,
                              family=binomial)
summary(xgb_false_model_role_cua)
# -------------------------------

dm_true_model_role_cua <- glmer(dm_correct ~ role + (1 | StudentID), data=dummy_predicted_gaming,
                               family=binomial)
summary(dm_true_model_role_cua)

dm_false_model_role_cua <- glmer(dm_correct ~ role + (1 | StudentID), data=dummy_predicted_not_gaming,
                                family=binomial)
summary(dm_false_model_role_cua)

```